Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-23 text/paper1/paper.tex
revisions

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision. \citet{madore2012}
have demonstrated \citep[using 5 stars with trigonometric parallaxes
  measured by Hubble;][]{benedict2011} that the dispersion in the
mid-IR PL relation -- at Spitzer wavelengths -- is of order
$\pm$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $\pm 2\%$ for RR
Lyrae stars out to $\sim$80~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times). Distance measurements of Blue Horizontal Branch and Blue Straggler stars at these distances in the halo typically achieve 15-25\% uncertainties. [can cite Alis Deason if necessary].

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission, Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy with $<$10\% distance errors for heliocentric distances up to $\sim$6~kpc for RR Lyrae stars.

@@

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} illustrates results from the LM10
N-body simulation of dwarf satellite disruption along the expected Sgr
orbit. The particle positions are shown in a coordinate system defined
by the orbit of the satellite orbit \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis ratios ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

@@

The prospect of combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris opens up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\sim$2~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim$10~km/s;][]{majewski2004}, and ground-based radial
velocities with $\sim$10~km/s accuracy, the measured phase-space
position of each individual RR Lyrae becomes a
powerful probe of the potential (see \ref{fig:}): the next section outlines a new
method to take advantage of this information.

@@

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi$. For a
given set of potential parameters, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=\Big(\frac{m_{\rm sat}}{3m_{\rm enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}

>>>>

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision. \citet{madore2012}
have demonstrated \citep[using five stars with trigonometric parallaxes
  measured by Hubble;][]{benedict2011} that the dispersion in the
mid-IR PL relation -- at Spitzer wavelengths -- is of order
$\sim$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $2\%$ for RR
Lyrae stars out to $\sim$80~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times). Distance measurements of Blue Horizontal Branch and Blue Straggler stars at these distances in the halo typically achieve $\sim$25\% uncertainties \citep[e.g.,][]{deason2012}.

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia2001}, planned for launch in late
2013, is an astrometric mission which aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
during the 5-year lifetime of the mission, Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy with $<$10\% distance errors for heliocentric distances up to $\sim$6~kpc for RR Lyrae stars.

@@

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} shows particle positions from the final time-step of an N-body simulation of dwarf satellite disruption along the expected Sgr
orbit in the best-fitting Milky Way halo model from LM10. [TODO: Maybe not true...] The positions are shown in a coordinate system defined
by the orbit of the satellite \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis ratios ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

@@

Combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris will open up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc ($\epsilon_{d}=0.6~{\rm kpc}$) is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\epsilon_{v}\sim2$~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim10$~km/s;][]{majewski2004}, and ground-based radial
velocities, the measured phase-space
position of each individual RR Lyrae becomes a
powerful probe of the potential (see \ref{fig:}). The next section outlines a new
method to take advantage of this information.

@@

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi({\boldsymbol\theta})$. For a
given set of potential parameters, $\boldsymbol\theta$, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become physically re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,;\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=\Big(\frac{m_{\rm sat}}{3m_{\rm enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}
